---
title: Reactomics analysis for MS-only data
author: Miao Yu
date: '2024-05-29'
slug: reactomics-analysis-for-ms-only-data
categories: []
tags: []
---

Recently, I received multiple requests of reactomics analysis for MS only data such as FT-ICR MS or MS imaging data. In this case, it's better to summary the answer with an example as reference. Here you are!

When retention time is not provided, m/z vector can still be used to check reaction level changes. To apply this analysis, you need to install the devel version(>=0.2.6) of PMD package:

```{r}
remotes::install_github('yufree/pmd')
```

You can still use `getrda` to find the high frequency PMDs. 

```{r}
library(pmd)
data(spmeinvivo)
# get the m/z
mz <- spmeinvivo$mz
# get the m/z intensity for all m/z, the row order is the same with mz
insms <- spmeinvivo$data
# check high frequency pmd
sda <- getrda(mz)
colnames(sda)
# save them as numeric vector
hfpmd <- as.numeric(colnames(sda))
```

Then `getpmddf` function can be used to extract all the paired ions for certain PMD.

```{r}
# get details for certain pmd
pmddf <- getpmddf(mz,pmd=18.011,digits = 3)
# add intensity for all the paired ions
mz1ins <- insms[match(pmddf$ms1,mz),]
mz2ins <- insms[match(pmddf$ms2,mz),]
# get the pmd pair intensity
pmdins <- mz1ins+mz2ins
# get the pmd total intensity across samples
pmdinsall <- apply(pmdins,2,sum)
# show the PMD intensity
pmdinsall
```

You can also calculate the static or dynamic PMD intensity for m/z only data.

```{r}
# get the ratio of larger m/z over smaller m/z
ratio <- mz2ins/mz1ins
# filter PMD based on RSD% across samples
# cutoff 30%
cutoff <- 0.3
# get index for static PMD
rsdidx <- apply(ratio,1,function(x) sd(x)/mean(x)<cutoff)
# get static PMD
pmddfstatic <- pmddf[rsdidx,]
# get static intensity
pmdinsstatic <- pmdins[rsdidx,]
# normalize the ions pair intensity to avoid influences from large response factors
pmdinsstaticscale <- t(scale(t(pmdinsstatic)))
# get the pmd static intensity across samples
pmdinsstaticall <- apply(pmdinsstaticscale,2,sum)
# show the PMD static intensity for each sample
pmdinsstaticall

# get index for dynamic PMD
rsdidx <- apply(ratio,1,function(x) sd(x)/mean(x)>=cutoff)
# get dynamic PMD
pmddfdynamic <- pmddf[rsdidx,]
# get dynamic intensity for ms1 and ms2
pmdinsdynamicms1 <- apply(mz1ins[rsdidx,],1,function(x) sd(x)/mean(x))
pmdinsdynamicms2 <- apply(mz2ins[rsdidx,],1,function(x) sd(x)/mean(x))
# find the stable ms and use ratio as intensity
idx <- pmdinsdynamicms1>pmdinsdynamicms2
pmdinsdynamic <- ratio[rsdidx,]
pmdinsdynamic[idx,] <- 1/ratio[rsdidx,][idx,]
# get the pmd dynamic intensity across samples
pmdinsdynamicall <- apply(pmdinsdynamic,2,sum)
# show the PMD dynamic intensity for each sample
pmdinsdynamicall
```

You can also use `getpmddf` function extract all the paired ions for multiple PMDs. Then you could generate the network based on the output.

```{r}
# get details for certain pmd
pmddf <- getpmddf(mz,pmd=hfpmd,digits = 3)
# viz by igraph package
library(igraph)
net <- graph_from_data_frame(pmddf,directed = F)
pal <- grDevices::rainbow(length(unique(E(net)$diff2)))
plot(net,vertex.label=NA,vertex.size = 5,edge.width = 3,edge.color = pal[as.numeric(as.factor(E(net)$diff2))],main = 'PMD network')
legend("topright",bty = "n",
       legend=unique(E(net)$diff2),
       fill=unique(pal[as.numeric(as.factor(E(net)$diff2))]), border=NA,horiz = F)
```

